Improving \textsl{Gaia} parallax precision \\ with a data-driven model of stars

### 2017-06-12 paper/paper.tex
paper figure updates

%@archiver{posteriorCMD.png}

@@

The \gaia\ Mission (\citealt{gaia16}) will soon deliver more than a billion
stellar parallax measurements.
Only a small fraction (but large number) of these
measurements will be both precise and purely astrometric.
\gaia\ will use astrometric parallaxes to determine the distances of the more
precise stars, and then calibrate spectrophotometric models.
These spectrophotometric models, in turn, along with \gaia's on-board
low-resolution $B_p\,R_p$ spectrophotometry, are used to provide
parallax estimates to less precise stars.
The full stack required for these parallax inferences is complex.
It involves modeling the stars, as well as the dust in the Milky Way,
and the response of the telescope itself.
It might be possible to eliminate some of this complexity using data driven models which don't rely on physical models.

@@

We now describe a singularity of our method, which allows us to conservatively deconvolve the data into a \cmd\ prior without requiring a full hierarchical inference framework (which was explored in \citealt{leistedtHogg2017}).
Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\data$ contains a transformation of the dust corrected absolute magnitude
\begin{equation}
y_0 = \varpi\,10^{\frac{1}{5}\,J^{\corr}_n} = 10^{\frac{1}{5}\,M^{\corr}_{J,n} + 2}
\label{eq:transform}
\end{equation}
where, for this expression, it is important to remember that $\varpi$ is assumed to be in units of $[mas]$.
The transformation of the dust corrected absolute magnitude is required to to keep the uncertainties Gaussian, which, under some assumptions along with this transformation, we maintain (Gaussian uncertainties are required to use \xd, see below).
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the 3D dust map. So our full terminology is
\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\truth_n = [10^{0.2\,M^{\true}_{J,n} + 2}, (J-K_s)^{\true}_n] \\
\mathrm{data \, vector} \quad &\data_n = [\varpi_n\,10^{0.2J^{\corr}_n}, (J-K_s)^{\corr}_n] \;\hspace*{15mm}  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^{\corr}_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J-K_s)^{\corr}_n = J_n - K_{s,n} - Q_{JK}\,E(B-V)_n \\
\mathrm{under\ the\ assumptions} \quad &I_n = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}
where $E(B-V)_n$ is the output of the dust model of \cite{green15}, and $Q_{\lambda}$ is the correction factor for a photometric band assuming some reddening law.

We assume that the parallax likelihood is Gaussian, as well as the $(J-K_s)^{\corr}_n$ color likelihood. So our full likelihood for the $n$th star is
\begin{equation}
p(\data_n \given \truth_n, I_n) = \mathcal{N}(\data_n \given \truth_n, \covar_n)
\end{equation}
with \\
\[
\covar_n = \begin{bmatrix}
\sigma_{J,n}^2 + \sigma_{K,n}^2 & 0 \\
0 & \left(\sigma_{\varpi}\,10^{0.2\,J_n^{\corr}}\right)^2
\end{bmatrix}
\]
where $\mathcal{N}(\mean, \mat{\Sigma})$ represents a normal distribution with
mean vector $\mean$ and covariance matrix $\mat{\Sigma}$.
Here we have assumed that the parallax uncertainty is significantly larger than the $J$-band magnitude uncertainty and is therefore the dominant uncertainty in the transformed absolute magnitude. We take $J_n^{\corr}$ as a point estimate and only propagate the parallax uncertainty.

@@

We use stars crossmatched in \tgas\ and \tmass.
The match was done using a nearest-neighbor algorithm with a search radius of $4"$.\footnote{\url{http://portal.nersc.gov/project/cosmo/temp/dstn/gaia/tgas-matched-2mass.fits.gz}}
We also required that the stars lie within the
observing footprint of \psone\ to access the \cite{green15} 3D dust model.
We require that the photometry have real values, and nonzero, real errors, and
remove a small selection of \tmass\ stars that have zero color and zero $J$-band
magnitude.
The full data set is visualized in the \cmd\ in \figurename~\ref{fig:data}.
The left panel shows the point estimates of the color and absolute magnitude,
using the point estimate of the parallax from \gaia, and the $1\sigma$ and
$2\sigma$ contours of the distribution.
The right panel shows a subset of the data with it's associated error bars, calculated via subsampling.
The uncertainties in the colors are fairly well behaved, but the large
uncertainties in some absolute magnitudes are due to the large uncertainties in
the parallax measurements.

@@

In detail, we take the $5\%$
quantile of each parallax posterior (the closest part) and query the 3D dust map at that distance.
We take the $5\%$ quantile---as apposed to the median---to do a minimal dust
correction because over-correcting for dust can push the likelihood
function into an unphysical, extremely blue part of the \cmd\ and thus bias the
parallax inference. At the 3D position of each star, we sample the \texttt{mode=median} value of the probabilistic dust map.
We apply each dust correction to our
\tmass\ photometry using Equation~\ref{eq:data}, where $Q_{\lambda}
= [0.709, 0.302]$ for bands $[J, K_s]$ respectively (see Table 6 in
\citealt{schlafly11}).
The dust values we obtain in our final
iteration, and which we apply to the photometry in our inference, are visualized in
galactic coordinates in \figurename~\ref{fig:dust}.
After applying the dust correction, we do not update the uncertainties, which is
technically wrong and under-represents the true uncertainties; this leads to a less severe deconvolution, as mentioned previously.
In other words, the features of our inferred
  \cmd\ are conservatively broad.\footnote{This can be understood by considering the following procedure: if one incrementally increased the noise levels (leaving the data unchanged; just increasing the declared uncertainties), the deconvolved distribution would be increasingly narrower, since it targets the distribution that would generate the noiseless observations, as well as the observed data once noise is added. Thus, under-estimating the uncertainties leads to a conservative under-deconvolution.}
We could add the covariance of
the dust to our likelihoods, but this is beyond the scope of this
demonstration of concept.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{dust.png}
\caption{The converged dust values at the 5\% distance quantile in Galactic coordinates. This shows both the footprint of our analysis as well as the adopted extinction values. Each point represents a single star in our cross-matched catalog. The large missing areas are due to the \psone\ footprint ($>-30~{\rm deg}$ in declination)}
\label{fig:dust}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{delta.png}
\caption{Fractional Change in Variance: The left panel shows the natural log of the fractional change in the variance of the posterior relative to the likelihood as a function of the stellar color. The coloring of the points in both panels is by the fractional change in variance to guide the coloring of the right panel. The blue lines are the $1\sigma$ and $2\sigma$ contours. The median fractional change is 1.2. Some posteriors have significantly smaller variances shown in black, but some actually increase in variance, shown as the yellow bump at $(J-K_s)^{\corr}$ of about 0.6. The right panel shows the posterior point estimates on the CMD colored by fraction change in variance. There is significant changes for the red giant branch stars, as well as some main sequence stars.}
\label{fig:delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{deltaCDF.png}
\caption{Fractional Change in Variance: The cumulative distribution function of the natural log of the fractional variance change when comparing our posterior to the \gaia\ catalogue. Negative numbers represent stars which have a posterior that is narrower than the \gaia\ catalogue, which is true for a large fraction of the stars. We've drawn two dashed lines. One represents the median star $(y=0.5)$ which has a fractional variation of $x=-0.3=ln(1/1.2^2)$, corresponding to an increase in precision by a factor of 1.2. Another represents the fraction of stars $(y=0.14)$ which have a fractional variance greater than $x=-1.4=ln(1/2^2)$, corresponding to an increase in precision by a factor of 2.}
\label{fig:deltaCDF}
\end{figure}

@@

We select sources from our catalog within a $1~{\rm deg}^2$ window centered on the position of M67; the cluster is visible as an over-density of stars in the sky positions alone.
To be clear, we do no prior selection on previously known members of M67, but include all stars observed by \gaia\ within this window on the sky.
Keep in mind, the (low) number of M67 members seen here is constrained by the (unknown) selection effects of the \tgas\ dataset.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
\figurename~\ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not at all obvious.
The posterior PDFs, shown in the bottom row, are more sharply peaked showing an increase in precision.
The posterior PDFs also have modes that are more similar to the previously determined distance to the cluster (vertical lines) for far more stars.
This shows that our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{distancesM67.png}
\caption{Proof of concept: parallax inference to the open cluster M67.
The upper left panel visualizes the \gaia\ + \tmass\ likelihoods of the data, shown in black and projected onto the \cmd, for all the stars within our $1~{\rm deg}^2$ window centered on the position of M67.
Our prior, visualized as a mixture of Gaussians, is shown in green.
The upper right panel shows the \gaia\ parallax likelihoods in black, with the ``true" distance to M67, determined by previous methods, bracketed by the red lines.
Similar to the upper left panel, the lower left panel shows the likelihoods of the data in black and our prior in green, as well as the posterior PDFs in blue.
The posterior PDFs are a mixture of Gaussians as well, each a product of the likelihood with a Gaussian in our prior.
The lower right panel shows the projection of the 2D posterior PDF onto parallax space.
The posterior PDFs are much more clustered and sharply peaked around the ``true" distance to M67 than the \gaia\ likelihoods, showing that our prior generates posterior PDFs that are accurate and more precise than \gaia\ alone.}
\label{fig:m67}
\end{figure}

>>>>

%@archiver{posteriorCMD.pdf}

@@

The \gaia\ Mission (\citealt{gaia16}) will soon deliver more than a billion
stellar parallax measurements.
Only a small fraction (but large number) of these
measurements will be both precise and purely astrometric.
\gaia\ will use astrometric parallaxes to determine the distances of the more
precise stars, and then calibrate spectrophotometric models.
These spectrophotometric models, in turn, along with \gaia's on-board
low-resolution $BP/RP$ spectrophotometry, are used to provide
parallax estimates to less precise stars.
The full stack required for these parallax inferences is complex.
It involves modeling the stars, as well as the dust in the Milky Way,
and the response of the telescope itself.
It might be possible to eliminate some of this complexity using data driven models which don't rely on physical models.

@@

We now describe a singularity of our method, which allows us to conservatively deconvolve the data into a \cmd\ prior without requiring a full hierarchical inference framework (which was explored in \citealt{leistedtHogg2017}).
Instead of transforming the parallax and apparent magnitude into an absolute magnitude, the data vector $\data$ contains a transformation of the dust corrected absolute magnitude
\begin{equation}
y_0 = \varpi\,10^{\frac{1}{5}\,J^{\corr}_n} = 10^{\frac{1}{5}\,M^{\corr}_{J,n} + 2}
\label{eq:transform}
\end{equation}
where, for this expression, it is important to remember that $\varpi$ is assumed to be in units of $[mas]$.
The transformation of the dust corrected absolute magnitude is required to keep the uncertainties Gaussian, and Gaussian uncertainties are required to use \xd, see below.
Our prior assumptions $I$ are the uncertainties in each of these measurements, as well as the 3D dust map. So our full terminology is
\begin{equation}
\begin{aligned}
\mathrm{true \, vector} \quad &\truth_n = [10^{0.2\,M^{\true}_{J,n} + 2}, (J-K_s)^{\true}_n] \\
\mathrm{data \, vector} \quad &\data_n = [\varpi_n\,10^{0.2J^{\corr}_n}, (J-K_s)^{\corr}_n] \;\hspace*{15mm}  n \in \mathrm{stars} \\
\mathrm{with \; apparent \; magnitude} \quad &J^{\corr}_n = J_n - Q_J\,E(B-V)_n \\
\mathrm{and \; color} \quad &(J-K_s)^{\corr}_n = J_n - K_{s,n} - Q_{JK}\,E(B-V)_n \\
\mathrm{under\ the\ assumptions} \quad &I_n = \left\{\left\{\sigma^2_{\varpi, n}\right\}, \left\{\sigma^2_{J,n}\right\}, \left\{\sigma^2_{K,n}\right\}, \mathrm{Dust \, Map}\right\}
\end{aligned}
\label{eq:data}
\end{equation}
where $E(B-V)_n$ is the output of the dust model of \cite{green15}, and $Q_{\lambda}$ is the correction factor for a photometric band assuming some reddening law.

We assume that the parallax likelihood is Gaussian, as well as the $(J-K_s)^{\corr}_n$ color likelihood. So our full likelihood for the $n$th star is
\begin{equation}
p(\data_n \given \truth_n, I_n) = \mathcal{N}(\data_n \given \truth_n, \covar_n)
\end{equation}
with \\
\[
\covar_n = \begin{bmatrix}
\left(\sigma_{\varpi}\,10^{0.2\,J_n^{\corr}}\right)^2 & 0 \\
0 & \sigma_{J,n}^2 + \sigma_{K,n}^2
\end{bmatrix}
\]
where $\mathcal{N}(\mean, \mat{\Sigma})$ represents a normal distribution with
mean vector $\mean$ and covariance matrix $\mat{\Sigma}$.
Here we have assumed that the parallax uncertainty is significantly larger than the $J$-band magnitude uncertainty and is therefore the dominant uncertainty in the transformed absolute magnitude. We take $J_n^{\corr}$ as a point estimate and only propagate the parallax uncertainty.

@@

We use stars crossmatched in \tgas\ and \tmass.
The match was done using a nearest-neighbor algorithm with a search radius of $4$ arcsec\footnote{\url{http://portal.nersc.gov/project/cosmo/temp/dstn/gaia/tgas-matched-2mass.fits.gz}}.
We also required that the stars lie within the
observing footprint of \psone\ to access the \cite{green15} 3D dust model.
We require that the photometry have real values, and nonzero, real errors, and
remove a small selection of \tmass\ stars that have zero color and zero $J$-band
magnitude.
The full data set is visualized in the \cmd\ in \figurename~\ref{fig:data}.
The left panel shows the point estimates of the color and absolute magnitude,
using the point estimate of the parallax from \gaia, and the $1\sigma$ and
$2\sigma$ contours of the distribution.
The right panel shows a subset of the data with its associated error bars, calculated via subsampling.
The uncertainties in the colors are fairly well behaved, but the large
uncertainties in some absolute magnitudes are due to the large uncertainties in
the parallax measurements.

@@

In detail, we take the $5\%$
quantile of each parallax posterior (the closest part) and query the 3D dust map at that distance.
We take the $5\%$ quantile---as apposed to the median---to do a minimal dust
correction because over-correcting for dust can push the likelihood
function into an unphysical, extremely blue part of the \cmd\ and thus bias the
parallax inference. At the 3D position of each star, we sample the \texttt{mode=median} value of the probabilistic dust map.
We apply each dust correction to our
\tmass\ photometry using Equation~\ref{eq:data}, where $Q_{\lambda}
= [0.709, 0.302]$ for bands $[J, K_s]$ respectively (see Table 6 in
\citealt{schlafly11}).
The dust values we obtain in our final
iteration, and which we apply to the photometry in our inference, are visualized in
galactic coordinates in \figurename~\ref{fig:dust}.
After applying the dust correction, we do not update the uncertainties, which is
technically wrong and under-represents the true uncertainties; this leads to a less severe deconvolution, as mentioned previously.
In other words, the features of our inferred
  \cmd\ are conservatively broad.\footnote{This can be understood by considering the following procedure: if one incrementally increased the noise levels (leaving the data unchanged; just increasing the declared uncertainties), the deconvolved distribution would be increasingly narrower, since it targets the distribution that would generate the noiseless observations, as well as the observed data once noise is added. Thus, under-estimating the uncertainties leads to a conservative under-deconvolution.}
We could add the covariance of
the dust to our likelihoods, but this is beyond the scope of this
demonstration of concept.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{dust.pdf}
\caption{The converged dust values at the 5\% distance quantile in Galactic coordinates. This shows both the footprint of our analysis as well as the adopted extinction values. Each point represents a single star in our cross-matched catalog. The large missing areas are due to the \psone\ footprint ($>-30~{\rm deg}$ in declination)}
\label{fig:dust}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{delta.pdf}
\caption{Fractional Change in Variance: The left panel shows the natural log of the fractional change in the variance of the posterior relative to the likelihood as a function of the stellar color. The coloring of the points in both panels is by the fractional change in variance to guide the coloring of the right panel. The blue lines are the $1\sigma$ and $2\sigma$ contours. The median fractional change is 1.2. Some posteriors have significantly smaller variances shown in black, but some actually increase in variance, shown as the yellow bump at $(J-K_s)^{\corr}$ of about 0.6. The right panel shows the posterior point estimates on the CMD colored by fraction change in variance. There is significant changes for the red giant branch stars, as well as some main sequence stars.}
\label{fig:delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{deltaParallax.pdf}
\caption{No bias in posterior pdf parallax expectation values.}
\label{fig:delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{deltaCDF.pdf}
\caption{Fractional Change in Variance: The cumulative distribution function of the natural log of the fractional variance change when comparing our posterior to the \gaia\ catalogue. Negative numbers represent stars which have a posterior that is narrower than the \gaia\ catalogue, which is true for a large fraction of the stars. We've drawn two dashed lines. One represents the median star $(y=0.5)$ which has a fractional variation of $x=-0.3=ln(1/1.2^2)$, corresponding to an increase in precision by a factor of 1.2. Another represents the fraction of stars $(y=0.14)$ which have a fractional variance greater than $x=-1.4=ln(1/2^2)$, corresponding to an increase in precision by a factor of 2.}
\label{fig:deltaCDF}
\end{figure}

@@

We select sources from our catalog within a $1~{\rm deg}^2$ window centered on the position of M67; the cluster is visible as an over-density of stars in the sky positions alone.
To be clear, we do no prior selection on previously known members of M67, but include all stars observed by \gaia\ within this window on the sky.
Keep in mind, the (low) number of M67 members seen here is constrained by the (unknown) selection effects of the \tgas\ dataset.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
\figurename~\ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not at all obvious.
The posterior PDFs, shown in the bottom row, are more sharply peaked showing an increase in precision.
The posterior PDFs also have modes that are more similar to the previously determined distance to the cluster (vertical lines) for far more stars.
This shows that our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{m67.pdf}
\caption{Proof of concept: parallax inference to the open cluster M67.
The upper left panel visualizes the \gaia\ + \tmass\ likelihoods of the data, shown in black and projected onto the \cmd, for all the stars within our $1~{\rm deg}^2$ window centered on the position of M67.
Our prior, visualized as a mixture of Gaussians, is shown in green.
The upper right panel shows the \gaia\ parallax likelihoods in black, with the ``true" distance to M67, determined by previous methods, bracketed by the red lines.
Similar to the upper left panel, the lower left panel shows the likelihoods of the data in black and our prior in green, as well as the posterior PDFs in blue.
The posterior PDFs are a mixture of Gaussians as well, each a product of the likelihood with a Gaussian in our prior.
The lower right panel shows the projection of the 2D posterior PDF onto parallax space.
The posterior PDFs are much more clustered and sharply peaked around the ``true" distance to M67 than the \gaia\ likelihoods, showing that our prior generates posterior PDFs that are accurate and more precise than \gaia\ alone.}
\label{fig:m67}
\end{figure}
